---
title: "StochOptim R Tests"
author: "Arqam Patel"
date: "2023-02-27"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)

```

## Rastrigin Function

Here, the Rastrigin function is implemented for n dimensions, with the default number of dimensions set to 2. Given an n-dimensional input vector $\mathbf{x} = (x_1, x_2, ..., x_n)$ , it outputs a scalar computed using the function:

$f(\mathbf{x}) = 2A + \sum_{i=1}^n [x_i^2 - A \cos(2\pi x_i)]$ where $A = 10$ and $n =2$ by default

```{r rastrigin}
rastrigin <- function(X, A = 10, n= 2){
  n*A + sum(X^2 - A*cos(2*pi*X))
}
```

## 3D Plot of the Rastrigin function

I use a perspective 3D plot to visualise the Rastrigin function in three dimensions, with different values on the x and y axes corresponding to $x_1$ and $x_2$ , and the z coordinate representing the function value $f(\mathbf{x})$ where $\mathbf{x} = (x_1, x_2)$


```{r echo = FALSE}

x <- seq(-5.12, 5.12, length.out = 100)
y <- seq(-5.12, 5.12, length.out = 100)
z <- matrix(0,100,100)

for(i in 1:length(x)){
  for(j in 1:length(y)){
    z[i,j] = rastrigin(c(x[i],y[j]))
  }
}

persp(x, y, z, theta = 30, phi = 30, expand = 0.5, main = "Rastrigin function 3D plot", cex.main =.8, col = "lightblue")
```

## Optimisation using optim()

Starting at some random coordinates, we try to find a local minimum of the above Rastrigin function surface using the optim optimiser.


```{r echo =F}

n <- 5

set.seed(1)
init <- matrix(runif(n = 2*n, -5.12, 5.12), nrow = n, ncol = 2)

vals <- numeric(n)
opt <- matrix(0, n, 3)

for(i in 1:n){
  vals[i] <- rastrigin(init[i,])
  optimum <- optim(par = init[i,], fn = rastrigin)
  opt[i, 1] <- optimum$par[1]
  opt[i, 2] <- optimum$par[2]
  opt[i, 3] <- optimum$value
}

mat <- cbind(init, vals, opt)
df <- as.data.frame(mat)
colnames(df) <- c("x_init", "y_init", "z_init", "x_optim", "y_optim", "z_optim")
df


```

We then vary the method for the same starting point:

```{r echo = F}
p <- init[1,]
mat <- as.data.frame(matrix(0,6,4))
mat[1,1:3] <- df[1,1:3]

j <-2

for(i in c("Nelder-Mead", "BFGS", "CG", "L-BFGS-B", "SANN"))
{
  optimum <- optim(par = p, fn = rastrigin, method = i)
  foo <- cbind(optimum$par[1], optimum$par[2], optimum$value, optimum$counts[1])
  mat[j,] <- foo
  j <- j+1
}

rownames(mat) <- c("initial","Nelder-Mead", "BFGS", "CG", "L-BFGS-B", "SANN")
colnames(mat) <- c("x", "y", "z", "f_evals") 
mat

```

optim() does not allow passing multiple methods as arguments at once, thus we have to use a loop.

## Other local optimisers

We use the ucminf optimiser to find minima starting from the exact same points as the previous implementation using optim(). 

```{r echo = F}
library(ucminf)
```

```{r echo = F}

vals2 <- numeric(n)
opt2 <- matrix(0, n, 3)

for(i in 1:n){
  vals2[i] <- rastrigin(init[i,])
  optimum <- ucminf(par = init[i,], fn = rastrigin)
  opt2[i, 1] <- optimum$par[1]
  opt2[i, 2] <- optimum$par[2]
  opt2[i, 3] <- optimum$value
}

mat2 <- cbind(init, vals2, opt2)
df2 <- as.data.frame(mat2)
colnames(df2) <- c("x_init", "y_init", "z_init", "x_optim", "y_optim", "z_optim")
df2

```

Alternatively, we can use optimx() to run multiple optimisers using a much smaller number of lines of code and compare them.

```{r echo = F}
library(optimx)
```

```{r echo =F}
methods <- c("Nelder-Mead", "BFGS", "CG", "L-BFGS-B","ucminf", "nlm", "nlminb")
optimx(par = init[1,] , rastrigin, method = methods)[, c(-7,-8,-9)]

```



## Global Optimisers

DEoptim, 

```{r echo = FALSE, include= F}
library(DEoptim)
library(GenSA)
library(rgenoud)
library(rbenchmark)
```

Now, we use multiple optimisation techniques and compare them in terms of time taken for 10 replications.

```{r include = F}

replications = 10
df3 <- as.data.frame(matrix(0, 3, 4), row.names = c("DEoptim", "GenSA", "genoud"))

df3[,4] <- benchmark(deoptim <- DEoptim(fn = rastrigin, lower = c(-5.12,-5.12), upper = c(5.12, 5.12), control = DEoptim.control(trace = F)), gensa <- GenSA(fn = rastrigin, lower = c(-5.12,-5.12), upper = c(5.12, 5.12)), rgenoud <- genoud(fn = rastrigin, nvars =2) , columns = "elapsed", replications = 10)

DEopt <- deoptim$optim

colnames(df3) = c("par1", "par2", "value", "time")

df3[1,1:3] <- c(DEopt$bestmem[1], DEopt$bestmem[2], DEopt$bestval) 
df3[2,1:3] <- c(gensa$par[1], gensa$par[2], gensa$value)
df3[3,1:3] <- c(rgenoud$par[1], rgenoud$par[2], rgenoud$value)

```

```{r echo = F}
df3
```

## Other functions


### Rosenbrock function

We implement the Rosenbrock function for n dimensions, and plot it for 2D inputs:

$f(\mathbf{x}) = \sum_{i=1}^{n-1} [ 100(x_{i+1} - x_i^2)^2 + (1-x_i)^2 ]$ 


```{r rosenbrock}
rosenbrock <- function(X, n = 2,  a = 1, b = 100){
  ans <- 0
  for(i in 1:(n-1)){
    ans <- ans + (a-X[i])^2 + b*(X[i+1] - X[i]^2)^2
  }
  return(ans)
}


```

```{r echo=FALSE}

x <- seq(-5.12, 5.12, length.out = 100)
y <- seq(-5.12, 5.12, length.out = 100)
z <- matrix(0,100,100)

for(i in 1:length(x)){
  for(j in 1:length(y)){
    z[i,j] = rosenbrock(c(x[i],y[j]))
  }
}

persp(x, y, z, theta = 30, phi = 30, expand = 0.5, main = "Rosenbrock function 3D plot", cex.main =.8, col = "lightblue")


```

### Styblinski–Tang function

We now implement the Styblinski-Tang function in n dimensions, given by:

$f(x) = \frac{1}{2} \sum_{i=1}^n (x_i^4 - 16 x_i^2 + 5x_i)$


```{r tang}

styb_tang <- function(X){
  sum(X^4 - 16*X^2 + 5*X)/2
}

```

```{r echo=FALSE}
# visualisation

x <- seq(-5, 5, length.out = 100)
y <- seq(-5, 5, length.out = 100)
z <- matrix(0,100,100)

for(i in 1:length(x)){
  for(j in 1:length(y)){
    z[i,j] = styb_tang(c(x[i],y[j]))
  }
}

persp(x, y, z, theta = 30, phi = 30, expand = 0.5, main = "ST function 3D plot", cex.main =.8, col = "lightblue")
```

## References
[CRAN Task View: Optimization and Mathematical Programming](https://cran.r-project.org/web/views/Optimization.html)

[Continuous Global Optimization in R](https://www.jstatsoft.org/article/view/v060i06)
[]()